\thetitle

### 2020-07-06 writing/challenges.tex
working on evaluation description + moderate reorganization

\section{Why is it hard to evaluate the value of blame?}

Creating a method for comparing blame assignment strategies poses several
challenges.

% \cd{One point we haven's justified is why empirical. The justification
% seems easy in the sense that we are looking at a real language, programs,
% bugs rather than models.}
%% MF: I edited the text to include this.

% \cd{I like this idea of the competent programmer. I couldn't have thought
% about homo economicus. Instead this was partly inspired by Gonen's work.}
% \cd{There they make a hypothesis, try to test it in vivo using designed model organisms
% with a well-specified first-order characteristic (e.g, a deleted gene).
% Then they collect data by placing the organism in a number of scenarios 
% and measuring aspects of its behavior and then they 
% construct an operational model from their data. Due to the synthetic
% nature of CS, for us the process goes in the opposite direction. In that
% sense the competent programmer is a model organism that, as its
% first-order characteristic, implements a direct consequence of the
% semantics. So in a sense for me this has always been about semantics and
% exploring their consequences.}

%% MF says: yes, I figured you were inspired by biology but in a
% synthetic world, like econ, you need also a rationale for why this
% thing is worthwhile measuring .. and so they came up with homo
% econ.

The most serious challenge concerns the nature of this research. Unlike ordinary
work on programming languages, the question seems to call for empirical research
similar to the one in the human-computer interaction area. At the same time, a
significant evaluation result demands a large number of cases yielding a serious
amount of data. But, as~\citet{lksfd-popl-2020} already demonstrated, the way
around this dilemma is to postulate the equivalent of a {\it homo economicus\/}
for programming languages---the {\em rational programmer\/}---and to implement
this rational programmer as an algorithm.  As indicated in the introduction, in
the context of gradual typing, the rational programmer translates the
Wadler-Findler slogan into a debugging method, searching for the source of the
impedance mismatch in an incremental fashion. Measuring this simulated
programmer's behavior on a large number of inputs is the equivalent of
collecting empirical data in a truly human-focused discipline.

One derivative challenge of the first one is the problem of simulating the
programmer's mistakes or, more precisely, the origin of impedance
mismatches. Before a programmer can try to find mistakes, mistakes have to be
made. Technically the problem is the lack of a sufficiently large available
corpus of type errors for a meaningful experiment. Similar to
~\citet{lksfd-popl-2020}, Following~\citet{lksfd-popl-2020} again, the solution
is to use mutation analysis~\cite{lipton1971fault, demillo1978hints,
jia2011analysis} to create mistakes in a synthetic manner. This process yields a
large number of buggy programs whose mixed-typed versions are suitable subjects
for our experiment.  The mutation analysis of Lazarek et al, is, however,
completely useless for creating impedance mismatches in a gradually typed
context.

While mutation analysis traditionally aims to inject bugs that challenge
extensive test suites, the ones yielding ill-typed mutants are usually deemed
\emph{incompetent}. As a matter of fact, mutation analysis frameworks are
fine-tuned to avoid them---yet it is precisely those that are needed for
evaluating type impedance mismatches. The ones used here are newly designed,
inspired by the authors extensive experience with programming in Typed Racket.
These type errors cover a wide range of Typed Racket's type system and moreover
reliably cause runtime exceptions in mixed-typed versions of mutants.

@@

The second big challenge is due to implementation disparity of gradual type
systems. While the various approaches are well understood in theory, they come
with vastly different type system, accommodate seemingly incompatible language
features, and apply to radically different coding styles in the untyped world.
To eliminate these differences as variables of a comparison, the application of
an evaluation method must use the same test programs and the same type system.
The case study presented here builds on Greenman and Felleisen's transient
variant for functional Typed Racket~\cite{gf-icfp-2018}. the revised transient
variant accommodates many more linguistic features and calls library code in a
way that understand the transient semantics.

Beyond the implementation issue, a comparison also faces the problem that the
three variants---the natural semantics, the transient semantics, and the erasure
semantics---come with three radically different blame assignment systems.  For
example, the natural semantics assigns blame to one component while transient
assigns blame to a set of components. The erasure semantics does not blame a
component per se, but it comes with an exception location and the usual stack
traces.

Making these strategies comparable is {\em not\/} an issue of language
implementation but one about how the rational programmer behaves with regard to
these messages. Specifically, the rational programmer comes with \emph{modes}
that represent the alternative ways in which a programmer may interpret blame
information. For instance, in transient, one mode has the competent programmer
pick the oldest element of a blame set as the next component to add types. After
all, this corresponds to the earliest point in the program's evaluation that can
discover the bug.  Another mode does not distinguish between the elements of the
blame set and requires that the rational programmer types all blamed components
at once.  Once information about the search steps in each mode is available, it
is normalized. In particular, it is not just the steps that are counted but the

Moreover, the three gradual typing systems differ in the type checks they
perform.  An extra mode aims to untangle the two issues. This \emph{vanilla} mode
ignores blame and instead uses information from dynamic type error messages to
select the next component to type. Hence it separates the effect of blame from
that of the type checks.

In sum, the peculiarities of the gradual typing landscape demand a careful
design of both the evaluation method and the application of the method to
specific gradual typing systems. Otherwise measuring the rational programmer
will yield potentially random results.

>>>>

\section{Why is it hard to evaluate the value of blame?}
\label{sec:challenges}

The most serious challenge for comparing blame assignment strategies concerns 
the nature of this research. Unlike ordinary
work on programming languages, the question seems to call for empirical research
similar to the one in the human-computer interaction area. At the same time, a
significant evaluation result demands a large number of cases yielding a serious
amount of data. But, as~\citet{lksfd-popl-2020} already demonstrated, the way
around this dilemma is to postulate the equivalent of a {\it homo economicus\/}
for programming languages---the {\em rational programmer\/}---and to implement
this rational programmer as an algorithm.  As indicated in the introduction, in
the context of gradual typing, the rational programmer translates the
Wadler-Findler slogan into a debugging method, searching for the source of the
impedance mismatch in an incremental fashion. Measuring this simulated
programmer's behavior on a large number of inputs is the equivalent of
collecting empirical data in a truly human-focused discipline.

One derivative challenge of the first one is the problem of simulating the
programmer's mistakes or, more precisely, the origin of impedance
mismatches. Before a programmer can try to find mistakes, mistakes have to
be made. Technically the problem is the lack of a sufficiently large
available corpus of type errors for a meaningful experiment.
Following~\citet{lksfd-popl-2020} again, the solution is to use mutation
analysis~\cite{lipton1971fault, demillo1978hints, jia2011analysis} to
create mistakes in a synthetic manner. This process yields a large number
of buggy programs whose mixed-typed versions are suitable subjects for our
experiment.  The mutation analysis of Lazarek et al, is, however,
completely useless for creating impedance mismatches in a gradually
typed context.

While mutation analysis traditionally aims to inject bugs that challenge
extensive test suites, the ones yielding ill-typed mutants are usually deemed
\emph{incompetent}. As a matter of fact, mutation analysis frameworks are
fine-tuned to avoid them---yet it is precisely those that are needed for
evaluating blame assignment strategies. The mutators used here are newly designed,
inspired by the authors extensive experience with programming in Typed Racket.
They cause type errors that cover a wide range of Typed Racket's type system and moreover
reliably cause runtime exceptions in mixed-typed versions of mutants.

@@

The second big challenge is due to the implementation disparity of gradual
type systems. While the various approaches are well understood in theory,
they come with vastly different type systems, accommodate seemingly
incompatible language features, and apply to radically different coding
styles in the untyped world.  To eliminate these differences as variables
of a comparison, the application of an evaluation method to different
gradual typing systems must use the same test programs and the same type
system. Otherwise, results for different gradual typing systems are not
comparable.  For instance, the case study presented here builds on
Greenman and Felleisen's transient variant for functional Typed
Racket~\cite{gf-icfp-2018} to obtain a transient semantics that is
comparable with Typed Racket's default natural one. The revised transient
variant accommodates many more linguistic features and calls library code
in a in a way that is faithful to the transient
semantics~\cite{vss-popl-2017}.

Beyond the implementation issue, a comparison also faces the problem that
the three variants---the natural semantics, the transient semantics, and
the erasure semantics---come with three radically different blame
assignment strategies.  For example, the natural semantics assigns blame
to one component while transient assigns blame to a set of components. The
erasure semantics does not blame a component per se, but it comes with an
exception location and the usual stack traces.

Making these strategies comparable is {\em not\/} an issue of language
implementation but one about how the rational programmer behaves with regard to
these messages. Specifically, the rational programmer must come with \emph{modes}
that represent the alternative ways in which a programmer may interpret blame. 
For instance, in transient, one mode has the competent programmer
pick the oldest element of a blame set as the next component to add types. After
all, this corresponds to the earliest point in the program's evaluation that can
discover an impedance mismatch.  Another mode does not distinguish between the elements of the
blame set and requires that the rational programmer types all blamed components
at once. Of course, a step of the rational programmer in the latter mode
is not as laborious as one in the former. Hence, an evaluation method should take 
into account the effort of the rational programmer when comparing 
different modes.

Besides the blame information they produce, the three gradual typing
systems also differ in the runtime type checks they perform. Therefore an
application of the evaluation method to a gradual typing system has to
factor in the effect of type checks to determine whether blame helps the
rational programmer find an impedance mismatch. A solution to this issue
is yet another mode for the rational programmer that uses information only
from teh underlying languages exceptions to select the next component(s) to type.
This \emph{vanilla} mode simulates a rational programmer that ignores
blame and serves as the baseline for blame's value.

The final big challenge is that the different gradual type
systems, the number of necessary modes for the rational programmer
and the large number of mutants lead to an explosion of scenarios the
method needs to examine. Thus a practical application of the
method has no option but to carefully sample this space while 
making sure that it still produces reproducible conclusions.

\smallskip

These challenges expose the subtlety of evaluating blame strategies but
also suggest a recipe for doing so correctly:

\begin{itemize}

\item Implement all the gradual type systems to be compared on a common
  basis (section~\ref{sec:landscape}).

\item Construct a large number of suitable benchmarks for all gradual type
  systems with representative
  type errors  (section~\ref{sec:mutate}).

\item Simulate the rational programmer with sufficient 
  modes to (i) match the peculiarities of each gradual type system and 
    (ii) isolate blame from other confounding factors
    (section~\ref{sec:rational}).

\item Sample the experimental space  to obtain a computationally feasible
  but statistically meaningful experiment (section~\ref{sec:sample}).

\end{itemize}

### 2020-07-06 writing/experiment.tex
titles are questions
working on evaluation description + moderate reorganization

\section{The Experimental Setup}

\begin{itemize}

\item Summary of the blame trail experiment. We sample and follow blame trails through configuration
        lattices.

\item Evaluating each point in the design space. We run the blame trail experiment for every point in the
        previously discussed design space.

\item Selection of programs. We are using a set of representative
  programs. Argue the programs are representative of the range of programs written by TR developers; point to the JFP paper.

\item Sampling of mutants. We have too many mutants to run them all, so we sample them with an aim to represent equally all of the mistakes captured by our operators.

>>>>

\subsection{How to Make Benchmarks for Blame}

\subsection{How to Sample Mutants and Configurations}

### 2020-07-06 writing/ideas.tex
fix basic latex mistakes in rudimentary way
working on evaluation description + moderate reorganization

\section{How to Evaluate Blame}

\begin{itemize}

    \item How to analyze this space?
          \begin{itemize}
            \item We want a quantative measure of error msg utility that allows
                  us to evaluate each point in the above space.
          \end{itemize}

     \item Blame trails
       \begin{itemize}

        \item We want an empirical study to understand how the behavior of blame
        in real systems.

        \item Blame trails provide a quantitative measure of a a system's
        error message utility, and they can be applied to every point
        in the above space.

       \item We generalize the blame trails idea of Lukas et al [POPL '20] to make it
        more systematic, scalable, and general, by:
           \begin{itemize}
             \item Adding a null hypothesis baseline for comparison between multiple systems
             \item Analyzing multiple blame following strategies
             \item Mutating interfaces as well as programs
             \item Incorporating domain knowledge in the selection of blame following
           strategies and mutation
             \item Evaluating systems on the same platform for apples-to-apples comparisons
           \end{itemize}

         \item Null hypothesis. We should have a baseline to tell how all of the points in the
        design space compare to having nothing at all.
\end{itemize}

>>>>

\section{How to Simulate the Rational Programmer} \label{sec:ideas}

Given the comparable implementations of the three gradual typing systems,
the next step for applying our method is to obtain a set of fully
typed benchmarks that each comes with a single impedance mismatch between
its type specifications and its code. As we briefly discuss in
section~\ref{sec:challenges}, we derive such ill-typed benchmarks with
mutation analysis from well-typed programs.  We return to mutation
analysis and the benchmarks in section~\ref{sec:experiment}.  Herein,
we assume that we have them in hand and instead we focus on the
main element of our method, the rational programmer.

The rational programmer exploits blame information to figure out wih. Therefore the way to understand its behavior 
is to analyze the migratory path it follows through the 
\emph{configuration space} of the program. The space
consists of variants of the program, dubbed \emph{configurations}, that 
differ in their type annotations. 
Herein we build on top of Typed Racket's migratory typing
approach,\footnote{The method can be adapted to so-called true gradual typing
systems~\cite{svcb-snapl-2015} in a straightforward manner.}, hence, each configuration differs from another
in which of its components are typed and which are not. Formally, we can
represent a program $\system$ as a set of of components
$\set{\component}$ and each configuration $\conf$ of $\system$ is a subset of 
$\set{\component}$
that contains exactly the components of  $\system$ that are typed in
$\conf$. Naturally the configurations of  $\system$ are ordered by the subset relation and
thus form a lattice $\lattice{\system}$ with size $2^{\size{\system}}$.
The bottom configuration of $\lattice{\system}$ is always $\emptyset$ and
the top element is $\system$ itself. In between there are all the
mixed-type configurations. If we assume that  component $\component$ of $\system$
contains the type error, any configuration $\conf$ that
includes $\component$ is ill-typed. Any other configuration $\conf$ may blame 
a $\component_b$, which we denote with $\blame{\conf}$. Let a blametrail
be a sequence of configurations $\conf_0,...\conf_n$ where 
for all $0 \leq i \leq n - 1$, $\conf_i \subset \conf_{i+1}$ and
$\conf_{i+1} \setminus \conf_i = \blame{\conf_i}$. That is a blametrail
corresponds to a succession of a configurations that start with a
configuration $\conf_0$, dubbed the \emph{root of the blametrail}, and 
each configuration is derived from its previous one by typing the
component that the latter blames. A rational programmer is one that
migrates the variant of a program $\system$ that corresponds to
configuration $\conf_0$ along the blametrail that starts at $\conf_0$.
We say a blametrail is final iff its last configuration, $\conf_n$
doesn't blame a component, $\blame{\conf_i} = \emptyset$. We call a
blametrail \emph{successful} iff it is final and its last configuration
$\conf_n$ contains the faulty component of $\system$, $\faulty{\system}
\in \conf_n$.

Using the notion of successful blametrails we can define precisely what it
means for blame to add value to a gradual typing system:

All blametrails are successfull: Given a program P and its configuration
lattice $\lattice{\system}$, all blametrails in the lattice are successful.

With these definitions we can describe with precision the experimental
process of our method. For each benchmark we create its configuration
lattice. Then we separate the configurations of the lattice that 
when run result in blame or result in a static type error from  those that do not
do either. The former are 
suitable to be roots of a blametrail. Thus we simulate the rational 
programmer and if the configuration blames we type their blamed component to obtain the next
configuration in the blametrail and repeat the process until the
blametrail becomes final. At this point we check whether the last
configuration fails to type check. If it does for all blametrails in the
lattice, then the property holds. Otherwise it fails.

In addition to the property the framework allows us to assign a metric of
effort to a successful blametrail; $\effort{\conf_0, ...., \conf_n} = n$

\subsection{How to Generalize the Rational Programmer}

As we discuss in section~\ref{sec:challenges}, an important factor that the method need to
take into account is that different gradual typing systems blame
differently and perform different checks. For that reason in
section~\ref{sec:challenges} we introduce the notion of modes for the
rational programmer. Defining these modes requires a generalization of the 
discussion above.

In particular it requires revisiting the notion of blametrail and that of
successful blametrail.

We define a blametrail in a mode  $\modem$ as  
a sequence of configurations $\conf_0,...\conf_n$ where 
for all $0 \leq i \leq n - 1$, $\conf_i \subset \conf_{i+1}$ and
$\conf_{i+1} \setminus \conf_i = \mode{\conf_i}$.

   For example we can obtain the simple blame mode
for the natural system if we use, as above, the selector $\blame{\cdot}$. 
However, the same mode has a different meaning in transient as this
semantics results in multiple blamed components.

This requires generalizing the notion of successful blametrail
to that of successful blametrees: a blametree is the set of all
blametrails with root a configuration $\conf_0$. A blametree is succesful if at least one
of its blametrails is succesful.

Of course for systems such as transient we can add modes where the rational
programmer picks one blametrail as part of the migration process. For
example the \first{\blame{\cdot}} mode continues with the oldest blame
entry in the blameset of a configuration, as we discuss in
section~\ref{sec:challenges}.

Moreover, the \emph{vanilla} mode corresponds to ignoring blame and instead
using only information from the dynamic type error that a configuration
produces using the $\error{\cdot}$ selector.

Finally, the \emph{null} mode uses selector $\random{\cdot}$ and it is
used by our method as the null hypothesis control experiment.

The following table collects all the modes and their selectors for the
three gradual typing systems we compare in this paper:

\begin{tabular}{c|c|c|c}
                    & {\bf Natural}  & {\bf Transient}        & {\bf Erasure} \\
   \hline 
  {\bf Simple Blame}   & \blame{\conf}  & \blame{\conf}          &               \\
 {\bf First Blame}  &                 & \first{\blame{\conf}}  &               \\
  {\bf Vanilla}       &  \error{\conf} & \error{\conf}          & \error{\conf} \\
  {\bf Null}        & \random{\conf} & \random{\conf}         & \random{\conf} 
\end{tabular}

As a last step we also need to lift the notion of effort to blametrees.
The effort of a blametree $\effort{\{\conf_0, ...., \conf_n\},...}$ is the
aggregate effort of all its blametrails.

### 2020-07-06 writing/introduction.tex
working on evaluation description + moderate reorganization

\section{Does Blame Matter?}

From the very beginning, the theory of gradual typing has come with a blame
theorem~\cite{mf-toplas-2009, tf-dls-2006}. ``Well-typed [components]\footnote{The
original authors got this word wrong.} can't be blamed'' turned the theorem into a
slogan~\cite{wf-esop-2009}. Indeed, academic implementationsa---for example,,
Reticulated Python~\cite{vsc-dls-2019, vss-popl-2017, vksb-dls-2014} and Typed
Racket~\cite{tf-dls-2006,tf-popl-2008,tfffgksst-snapl-2017,tf-icfp-2010}---come
with sophisticated blame-assignment subsystems. Theoreticians firmly embrace the
idea that blame can help practicing programmers find impedance mismatches between
the types they added to a software system and the behavior remaining untyped
components.

Practical gradual type systems ignore blame assignment almost completely.
Industrial systems such as Flow,\footnote{flow} Hack,\footnote{hack} or
TypeScript~\cite{ts} exploit types for IDE actions and for finding typos in
code. Then these compilers remove types and rely on the built-in safety checks of
the underlying programming language to catch any problems.

The contrast between academic and industrial systems raises the question whether
\begin{quote}
 \it
 blame assignment adds any value to a gradually typed language, \\
 especially for the benefit of the working programmer.
\end{quote}
Given the prominence of blame in academia and its complete absence in
industrial systems, it comes as an even bigger surprise that the research
literature and the industrial blog world do not present or discuss any possible
answers.  Instead, language designers seem to answer this question one way or
another without any scientific justification, and even when they have a language,
they neglect to look into this question. The research community simply does not
have a method for evaluating such questions, and given the question's practical
relevance, this is a problem.

This paper's first contribution is an automated method for evaluating the
effectiveness of blame assignment strategies in the gradual typing world (see
section~\ref{sec:ideas}). Roughly speaking, the method simulates a programmer who
follows the Wadler-Findler slogan quoted above. Its implied method suggests that
programmers can find impedance-mismatch bugs by adding types to a partially typed
software system. That is, when the run-time system signals an impedance mismatch,
a good blame messages helps programmers infer where to add an additional type
specification. Since components that pass the type checker cannot cause such
impedance signals; the type checker must discover these mismatches at compile
time.  If the additional type does {\em not\/} cause a static type error, the
program is run again until it signals another violation.  Counting the number of
steps it takes to discover the mistake is a measure of the effectiveness of the
blame assignment.  The method is loosely based on the work of Lazarek et
al.~\cite{lksfd-popl-2020}.

The paper's second contribution is a set of results from applying the evaluation
method to three distinct blame assignment systems and approximately 20,000
cases\mf{Replace with real number} (see section~\ref{sec:experiment}): Vitousek et
al.'s strategy of tracking typed/untyped boundaries in Reticulated;
Tobin-Hochstadt et al.'s use of higher-order contract system~\cite{ff-icfp-2002,
mf-toplas-2009}; and the error messages from failed safety checks in the
underlying programming language.

@@

The results of the experiment (see section~\ref{sec:results}) validate the
conjecture behind the work of theoreticians. A good blame assignment system
greatly shortens the search for impedance mismatches between the specified types
and the behavior of untyped components. Even better, the evaluation method clearly
identifies the wrapper-based blame tracking of Typed Racket as the vastly superior
academic system. Vitousek et al. lose one more time. Of course, a detailed
discussion must acknowledge the threats to validity of these results (see
section~\ref{sec:discussion}).

Next, the paper provide an overview of the blame assignment strategies as found in
existing gradually typed language implementation (see section~\ref{sec:landscape})
and the challenges of developing a systematic evaluation method
(section~\ref{sec:challengs}).

>>>>

\section{Does Blame Matter?} \label{sec:introduction}

From the very beginning, the theory of gradual typing has come with a blame
theorem~\cite{mf-toplas-2009, tf-dls-2006}. ``Well-typed [components]\footnote{The
original authors got this word wrong.} can't be blamed'' turned the theorem into a
slogan~\cite{wf-esop-2009}. Indeed, academic implementations---for example,
Reticulated Python~\cite{vsc-dls-2019, vss-popl-2017, vksb-dls-2014} and Typed
Racket~\cite{tf-dls-2006,tf-popl-2008,tfffgksst-snapl-2017,tf-icfp-2010}---come
with sophisticated blame assignment strategies. Theoreticians firmly embrace the
idea that blame can help practicing programmers find impedance mismatches between
the types they added to a software system and the behavior of the remaining untyped
components.

Practical gradual type systems ignore blame assignment almost completely.
Industrial systems such as Flow,\footnote{\url{https://flow.org}}
Hack,\footnote{\url{https://hacklang.org}} or
TypeScript\footnote{\url{https://www.typescriptlang.org}} exploit types for IDE actions and for finding typos in
code. Then these compilers remove types and rely on the built-in safety checks of
the underlying programming language to catch any problems.

The contrast between academic and industrial systems raises the question whether
\begin{quote}
 \it
 blame assignment adds any value to a gradually typed language, \\
 especially for the benefit of the working programmer.
\end{quote}
Given the prominence of blame in academia and its complete absence in
industrial systems, it comes as an even bigger surprise that the research
literature and the industrial blog world do not present or discuss any
possible answers.  Instead, when language designers decide to look into the
question, they seem to answer it one way or another without any scientific
justification. Most of the time though they ignore the question altogether. The
root of the issue is that the research community simply does not have a
method for evaluating such questions, and given the question's practical
relevance, this a significant problem.

This paper's first contribution is an automated method for evaluating the
effectiveness of blame assignment strategies in the gradual typing world
(sections ~\ref{sec:landscape},~\ref{sec:mutate},~\ref{sec:rational} and \ref{sec:sample}). 
Roughly speaking, the method, which is
loosely based on the work of Lazarek et al.~\cite{lksfd-popl-2020},
simulates a programmer who follows the Wadler-Findler slogan quoted above.
Such a \emph{rational programmer} can find impedance-mismatch
bugs by adding types to a partially typed program. That is, when the
run-time system signals an impedance mismatch, a good blame message helps
the programmer infer where to add an additional type specification. Since
components that pass the type checker cannot cause impedance mismatches;
the type checker discovers these mismatches at compile time and points
directly to the faulty component. Otherwise, if the
additional type does {\em not\/} cause a static type error, the program
 is run again to signal another violation. Then the programmer can heed
blame one more time to add yet another type specification. Counting the number of
steps it takes to discover the mistake in this manner is a measure of the effectiveness
of the blame assignment strategy of the language.

The paper's second contribution is a set of results from applying the evaluation
method to three distinct blame assignment strategies and approximately 20,000
cases\mf{Replace with real number} (section~\ref{sec:results}): Vitousek et
al.'s strategy of tracking typed/untyped boundaries in Reticulated;
Tobin-Hochstadt et al.'s use of higher-order contract system~\cite{ff-icfp-2002,
mf-toplas-2009}; and the approach of industrial systems that forego blame
and instead rely on error messages from the
underlying programming language.

@@

The results of the experiment (section~\ref{sec:discussion}) validate the
conjecture behind the work of theoreticians. A good blame assignment
strategy
greatly shortens the search for impedance mismatches between the specified types
and the behavior of untyped components. Even better, the evaluation method clearly
identifies the wrapper-based blame tracking of Typed Racket as the vastly superior
academic system. Vitousek et al. lose one more time. Of course, a detailed
discussion must acknowledge the threats to validity of these results.

Next, the paper describes the challenges of developing an evaluation method
(section~\ref{sec:challenges}), which may explain why the question has been neglected
for such a long time.

### 2020-07-06 writing/mutate.tex
working on evaluation description + moderate reorganization

\section{How to Simulate the Rational Programmer} \label{sec:ideas}

Given the comparable implementations of the three gradual typing systems,
the next step for applying our method is to obtain a set of fully
typed benchmarks that each comes with a single impedance mismatch between
its type specifications and its code. As we briefly discuss in
section~\ref{sec:challenges}, we derive such ill-typed benchmarks with
mutation analysis from well-typed programs.  We return to mutation
analysis and the benchmarks in section~\ref{sec:experiment}.  Herein,
we assume that we have them in hand and instead we focus on the
main element of our method, the rational programmer.

The rational programmer exploits blame information to figure out wih. Therefore the way to understand its behavior 
is to analyze the migratory path it follows through the 
\emph{configuration space} of the program. The space
consists of variants of the program, dubbed \emph{configurations}, that 
differ in their type annotations. 
Herein we build on top of Typed Racket's migratory typing
approach,\footnote{The method can be adapted to so-called true gradual typing
systems~\cite{svcb-snapl-2015} in a straightforward manner.}, hence, each configuration differs from another
in which of its components are typed and which are not. Formally, we can
represent a program $\system$ as a set of of components
$\set{\component}$ and each configuration $\conf$ of $\system$ is a subset of 
$\set{\component}$
that contains exactly the components of  $\system$ that are typed in
$\conf$. Naturally the configurations of  $\system$ are ordered by the subset relation and
thus form a lattice $\lattice{\system}$ with size $2^{\size{\system}}$.
The bottom configuration of $\lattice{\system}$ is always $\emptyset$ and
the top element is $\system$ itself. In between there are all the
mixed-type configurations. If we assume that  component $\component$ of $\system$
contains the type error, any configuration $\conf$ that
includes $\component$ is ill-typed. Any other configuration $\conf$ may blame 
a $\component_b$, which we denote with $\blame{\conf}$. Let a blametrail
be a sequence of configurations $\conf_0,...\conf_n$ where 
for all $0 \leq i \leq n - 1$, $\conf_i \subset \conf_{i+1}$ and
$\conf_{i+1} \setminus \conf_i = \blame{\conf_i}$. That is a blametrail
corresponds to a succession of a configurations that start with a
configuration $\conf_0$, dubbed the \emph{root of the blametrail}, and 
each configuration is derived from its previous one by typing the
component that the latter blames. A rational programmer is one that
migrates the variant of a program $\system$ that corresponds to
configuration $\conf_0$ along the blametrail that starts at $\conf_0$.
We say a blametrail is final iff its last configuration, $\conf_n$
doesn't blame a component, $\blame{\conf_i} = \emptyset$. We call a
blametrail \emph{successful} iff it is final and its last configuration
$\conf_n$ contains the faulty component of $\system$, $\faulty{\system}
\in \conf_n$.

Using the notion of successful blametrails we can define precisely what it
means for blame to add value to a gradual typing system:

All blametrails are successfull: Given a program P and its configuration
lattice $\lattice{\system}$, all blametrails in the lattice are successful.

With these definitions we can describe with precision the experimental
process of our method. For each benchmark we create its configuration
lattice. Then we separate the configurations of the lattice that 
when run result in blame or result in a static type error from  those that do not
do either. The former are 
suitable to be roots of a blametrail. Thus we simulate the rational 
programmer and if the configuration blames we type their blamed component to obtain the next
configuration in the blametrail and repeat the process until the
blametrail becomes final. At this point we check whether the last
configuration fails to type check. If it does for all blametrails in the
lattice, then the property holds. Otherwise it fails.

In addition to the property the framework allows us to assign a metric of
effort to a successful blametrail; $\effort{\conf_0, ...., \conf_n} = n$

\subsection{How to Generalize the Rational Programmer}

As we discuss in section~\ref{sec:challenges}, an important factor that the method need to
take into account is that different gradual typing systems blame
differently and perform different checks. For that reason in
section~\ref{sec:challenges} we introduce the notion of modes for the
rational programmer. Defining these modes requires a generalization of the 
discussion above.

In particular it requires revisiting the notion of blametrail and that of
successful blametrail.

We define a blametrail in a mode  $\modem$ as  
a sequence of configurations $\conf_0,...\conf_n$ where 
for all $0 \leq i \leq n - 1$, $\conf_i \subset \conf_{i+1}$ and
$\conf_{i+1} \setminus \conf_i = \mode{\conf_i}$.

   For example we can obtain the simple blame mode
for the natural system if we use, as above, the selector $\blame{\cdot}$. 
However, the same mode has a different meaning in transient as this
semantics results in multiple blamed components.

This requires generalizing the notion of successful blametrail
to that of successful blametrees: a blametree is the set of all
blametrails with root a configuration $\conf_0$. A blametree is succesful if at least one
of its blametrails is succesful.

Of course for systems such as transient we can add modes where the rational
programmer picks one blametrail as part of the migration process. For
example the \first{\blame{\cdot}} mode continues with the oldest blame
entry in the blameset of a configuration, as we discuss in
section~\ref{sec:challenges}.

Moreover, the \emph{vanilla} mode corresponds to ignoring blame and instead
using only information from the dynamic type error that a configuration
produces using the $\error{\cdot}$ selector.

Finally, the \emph{null} mode uses selector $\random{\cdot}$ and it is
used by our method as the null hypothesis control experiment.

The following table collects all the modes and their selectors for the
three gradual typing systems we compare in this paper:

\begin{tabular}{c|c|c|c}
                    & {\bf Natural}  & {\bf Transient}        & {\bf Erasure} \\
   \hline 
  {\bf Simple Blame}   & \blame{\conf}  & \blame{\conf}          &               \\
 {\bf First Blame}  &                 & \first{\blame{\conf}}  &               \\
  {\bf Vanilla}       &  \error{\conf} & \error{\conf}          & \error{\conf} \\
  {\bf Null}        & \random{\conf} & \random{\conf}         & \random{\conf} 
\end{tabular}

As a last step we also need to lift the notion of effort to blametrees.
The effort of a blametree $\effort{\{\conf_0, ...., \conf_n\},...}$ is the
aggregate effort of all its blametrails.

>>>>

\section{How to Make Benchmarks} \label{sec:mutate}

Given the comparable implementations of the three gradual typing systems,
the next step for applying our method is to obtain a set of fully
typed benchmarks that each comes with a single impedance mismatch between
its type specifications and its code. As we briefly discuss in
section~\ref{sec:challenges}, we derive such ill-typed benchmarks with
mutation analysis from well-typed programs.

### 2020-07-06 writing/paper.tex
challenges first
working on evaluation description + moderate reorganization

%% Abstract
%% Note: \begin{abstract}...\end{abstract} environment must come
%% before \maketitle command
\begin{abstract}
Programming language theoreticians develop blame assignment systems
and prove blame theorems for gradually typed programming languages.
Practical implementations of gradual typing almost completely ignore
the idea of blame assignment.  This contrast raises the question
whether blame assignment provides any value to the working programmer
and poses the challenge of how to evaluate the effectiveness of blame
assignment systems. This paper contributes (1) the first evaluation
method for blame assignment systems and (2) the results from applying
it to three different systems. \mf{This sentence must be checked
before final submission.} These results confirm that blame assignment
is helpful to a rationally acting programmer.
\end{abstract}

@@

\input{introduction.tex} \label{sec:introduction}
\input{landscape.tex}    \label{sec:landscape}
\input{challenges.tex}   \label{sec:challengs}
\input{ideas.tex}        \label{sec:ideas}
\input{experiment.tex}   \label{sec:experiment}
\input{results.tex}      \label{sec:results}
\input{discussion.tex}   \label{sec:discussion}
\input{related.tex}      \label{sec:related}
\input{conclusion.tex}   \label{sec:conclusion}

>>>>

%% Abstract
%% Note: \begin{abstract}...\end{abstract} environment must come
%% before \maketitle command
\begin{abstract}
Programming language theoreticians develop blame assignment systems
and prove blame theorems for gradually typed programming languages.
Practical implementations of gradual typing almost completely ignore
the idea of blame assignment.  This contrast raises the question
whether blame assignment provides any value to the working programmer
and poses the challenge of how to evaluate the effectiveness of blame
assignment strategies. This paper contributes (1) the first evaluation
method for blame assignment strategies and (2) the results from applying
it to three different gradual type systems. \mf{This sentence must be checked
before final submission.} These results confirm that blame assignment
is helpful to a rationally acting programmer.
\end{abstract}

@@

\input{introduction.tex} 
\input{challenges.tex}   
\input{landscape.tex} 
\input{mutate.tex}
\input{rational.tex}
\input{sample.tex}
\input{results.tex}      \label{sec:results}
\input{discussion.tex}   \label{sec:discussion}
\input{related.tex}      \label{sec:related}
\input{conclusion.tex}   \label{sec:conclusion}
